Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-24 pnas/Towards.tex
introduction pnas

\dropcap{C}ryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules. \TODO{revise}
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2-D tomographic image of the samples embedded in the ice, called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

The signal to noise ratio ($\SNR$) of the projections in the micrographs is a function of two dominating factors. On the one hand, the $\SNR$ is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the $\SNR$ is a function of the molecule \TODO{size/weight}. The smaller the molecules, the fewer detected electrons carry information about them.

All contemporary methods in the field split the reconstruction procedure into several stages.
The first stage consists in extracting the various particle projections from the micrographs. This stage is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. Figure~\ref{fig:micro_example} illustrates how particle picking becomes increasingly challenging as the $\SNR$ degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$; see for instance~\cite{bendory2018estimation}\TODO{To verify the title}. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2\times $  Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
\TODO{As these two leaders of the cryo-EM community point out,} it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs.

@@

As a proof of concept, we study a simplified model for 3-D reconstruction directly from the micrograph on a simulated data. 
In order to recover the 3-D volume, we use autocorrelation analysis. In a nutshell, we relate the autocorrelations of the micrographs to the autocorrelations of the volume.
For any noise level, these autocorrelations can be estimated to any desired accuracy, provided the we get to observe sufficiently many particle projections and the latter are separated in the micrograph. Importantly, there is no need to detect the projections. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the volume. To estimate the volume itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares. In this controlled experiment, we make several simplified assumptions; see the Discussion Section for details. 
\TODO{Do we want to stress that we get only low-resolution of the volume?}

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,vanheel1992correlation,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using template matching and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates or human intervention are required, thus significantly reducing concerns about model bias. \TODO{To add reference to our example.}

To illustrate the possibility to estimate a signal without detecting its locations, we first discuss a toy model for which it is easier to convey the mathematical principles. 
In this toy model, an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~TKTK for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the $\SNR$ is low enough that particle picking cannot be done reliably. 
This problem is interesting of its own as it appears in other scientific applications, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}. We study it in more details, including an extension to multiple planted images, in a companion paper~\cite{bendory2018estimation}.

@@

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done. \TODO{Do we want to replace with a cryo-EM figure? This figure will be positioned with the rest of the figures about Einstein's example}}	
\end{figure*}

@@

If $K=1$ and $x$ is known, then the locations $s_i$ can be estimated via linear programming  in the high $\SNR$ regime~\cite{azais2015spike,denoyelle2017support,bendory2016robust,bendory2017robust,bernstein2017deconvolution}. However, in the low $\SNR$ regime, estimating the binary sparse signal $s$ is impossible. We have few evidences for that\cite{aguerrebere2016fundamental}, but we will provide Will's argument here.
%
%To see this, suppose that an oracle provides us $M$ windows of length $W>L$, each containing exactly one copy of $x$. Suppose too that the oracle provides us with $x$ itself. That is to say, we get a series of windows of length $W$, each one containing a signal $x$ at an unknown location; and our only task is to estimate the locations. This is an easier problem than detecting the support of $s$. Nevertheless, even this simpler problem is impossible in the low $\SNR$ regime~\cite{aguerrebere2016fundamental}. Consequently, detecting the nonzero values of $s$ is impossible in low $\SNR$.

%\TODO{Here we should give the Will's proposition for the impossibility of detection}

@@

%%%Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low $\SNR$ regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

First, the second result of Proposition~\ref{prop:uniqueness} shows that there exists a very simple estimator that has finite sensitivity. In the next numerical experiments we use an estimator based on nonconvex LS that shows empirical robustness to additive noise, in accordance with related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous}. 
Second, these results carry through to signals of any dimension.
Third, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations and 
therefore the assumption that the first and last entries are nonzero is met. In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low $\SNR$ regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

@@

A few remarks are in order. 
First, the second result of Proposition~\ref{prop:uniqueness} shows that there exists a very simple estimator that has finite sensitivity. In the next numerical experiments we use an estimator based on nonconvex LS that shows empirical robustness to additive noise, in accordance with related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous}. 
Second, these results carry through to signals of any dimension.
Third, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations and 
therefore the assumption that the first and last entries are nonzero is met. In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low $\SNR$ regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

@@

In the  experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to $\SNR=1/20$). This high noise level is illustrated in TKTK. %In this first experiment, we assume knowledge of $\sigma$ and of the total number of signal occurrences across all micrographs.\TODO{Need to mention that in the 3-D reconstruction we do not make these assumptions.}
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction. The separation restriction is discussed in more details in the Methods Section.

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{Einstein_progress_examples}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
		%\TODO{To add: redo the figures according to Amit's comments and the figures of noisy Einstein's micrographs}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure*}

@@

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable $\SNR$ is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might.

>>>>

\dropcap{C}ryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules. \TODO{revise first sentence}
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. %Within the ice, the molecules are randomly oriented and positioned.
 The microscope produces a 2-D tomographic image of the samples embedded in the ice, called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

The signal to noise ratio (SNR) of the projections in the micrographs is a function of two dominating factors. On the one hand, the SNR is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the SNR is a function of the molecule \TODO{size/weight}. The smaller the molecules, the fewer detected electrons carry information about them.

All contemporary methods in the field split the reconstruction procedure into several stages.
The first stage consists in extracting the various particle projections from the micrographs. This stage is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. Figure~\ref{fig:micro_example} illustrates how particle picking becomes increasingly challenging as the SNR degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical SNR \TODO{the impossibility result should be in the theory section}. This fact has been recognized early on by the cryo-EM community. In particular, in an influential paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
Joachim Frank voices a similar observation in his Nobel prize lecture: ``\emph{Using the ribosome as an example, it became clear from the formula we obtained that the single-particle approach to structure research was indeed feasible for molecules of sufficient size: Particle Size > 3/[Contrast$^2\times $  Resolution (as length) $\times$ Critical Electron Dose]}''~\cite{frank2018single}. 
\TODO{As these two leaders of the cryo-EM community point out,} it is impossible to reconstruct small molecules by any of the existing computational pipelines for single particle analysis in cryo--EM, because the particles themselves cannot be picked from the micrographs.

@@

As a proof of concept, we study a toy model for which it is easier to convey the mathematical principles. 
In this toy model, an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the SNR is low enough that particle picking cannot be done reliably. 
This problem is interesting of its own as it appears in other scientific applications, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

In order to recover the image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelations of the micrographs to the autocorrelations of the image.
For any noise level, these autocorrelations can be estimated to any desired accuracy, provided that we get to observe sufficiently many image occurrences and the latter are separated in the micrograph. Importantly, there is no need to detect individual image occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem.

Later, we derive the mathematical relation between the autocorrelations of the micrographs and the 3-D volume under some simplifying conditions. We show a few numerical examples
 and draw the future developments required to make this method applicable on real data. \TODO{This paragraph would be changed according to the final results we will show}

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,vanheel1992correlation,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using template matching and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates or human intervention are required, thus significantly reducing concerns about model bias. %
%To illustrate the possibility to estimate a signal without detecting its locations, we first discuss a toy model for which it is easier to convey the mathematical principles. 
%In this toy model, an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~TKTK for an illustration.
%The goal is to estimate the planted image. The task is interesting in particular when the SNR is low enough that particle picking cannot be done reliably. 
%This problem is interesting of its own as it appears in other scientific applications, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}. We study it in more details, including an extension to multiple planted images, in a companion paper~\cite{bendory2018estimation}.

@@

\begin{figure*}[h]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contains the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low SNR, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure*}

@@

If $K=1$ and $x$ is known, then the locations $s_i$ can be estimated via linear programming  in the high SNR regime~\cite{azais2015spike,denoyelle2017support,bendory2016robust,bendory2017robust,bernstein2017deconvolution}. However, in the low SNR regime, estimating the binary sparse signal $s$ is impossible. We have few evidences for that\cite{aguerrebere2016fundamental}, but we will provide Will's argument here.

@@

%%%Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low SNR regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

First, the second result of Proposition~\ref{prop:uniqueness} shows that there exists a very simple estimator that has finite sensitivity. In the next numerical experiments we use an estimator based on nonconvex LS that shows empirical robustness to additive noise, in accordance with related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous}. 
Second, these results carry through to signals of any dimension.
Third, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations and 
therefore the assumption that the first and last entries are nonzero is met. In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low SNR regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

@@

A few remarks are in order. 
First, the second result of Proposition~\ref{prop:uniqueness} shows that there exists a very simple estimator that has finite sensitivity. In the next numerical experiments we use an estimator based on nonconvex LS that shows empirical robustness to additive noise, in accordance with related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous}. 
Second, these results carry through to signals of any dimension.
Third, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations and 
therefore the assumption that the first and last entries are nonzero is met. In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Finally, computing the $d$th autocorrelation amplifies the variance of the noise by a factor $d$ in the low SNR regime. Therefore, if we can estimate $a_z^3$ up to small perturbation, it implies that we can estimate $a_z^2$ accurately as the proposition assumes.

@@

In the  experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$=1/20$). This high noise level is illustrated in TKTK. %In this first experiment, we assume knowledge of $\sigma$ and of the total number of signal occurrences across all micrographs.\TODO{Need to mention that in the 3-D reconstruction we do not make these assumptions.}
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction. The separation restriction is discussed in more details in the Methods Section.

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{Einstein_progress_examples}
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.
		%\TODO{To add: redo the figures according to Amit's comments and the figures of noisy Einstein's micrographs}
		% At a noise level of $\sigma = 3$, this amounts to an SNR of $1/20$.
	}
\end{figure*}

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{1D_example}
	\caption{\label{fig:1D_example} Recovery of a 1-D signal of length $L=21$ at noise level $\sigma = 3$ from $M$ occurances in $N = 410*M$. he true value of $gamma$ is $0.0512$.
		  The first three panels show reconstruction with different $M$ values compared to the ground truth signal (in blue). The relative error of $\gamma$ for the three panels is: $27.1\%,4\%,1.2\%$. The right panel shows the relative error as a function of $M$.)
	}
\end{figure*}

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable SNR is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might.
